Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-20 cordillera-cycle.tex
Rework conclusions.

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Fraser Plateau lineation system predates
deglaciation, and that deglacial landforms are largely absent. Alternatively,
local geothermal heat associated with volcanic activity on the Fraser Plateau
could have triggered the basal sliding \citep{Fahnestock.etal.2001}.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glacial maxima during
MIS~4 (61.9--55.4\,ka) and MIS~2 (29.5--16.9\,ka), two periods
corresponding to documented extensive glaciations. This result is
independent of the choice of the palaeo-temperature record used to approximate
the past climate evolution, and thus
can be seen as a first-order agreement between the model and the geological
evidence. However, the timing of the two glaciation peaks depends sensitively
on which record
is used to drive the model. The timing of the LGM is best
reproduced by the Antarctic record, and occurs too early in all simulations
that are driven by the other records. The mismatch is greatest when using
oceanic records from the Pacific Northwest, which are affected by the
weakening of the California current during the LGM.

In all simulations presented here, ice cover is limited to disjoint mountain ice
caps during most of the glacial cycle, confirming previous inferences from the
geological evidence preceding the LGM. However, our
simulations produce persistent ice cover on the Skeena Mountains during
the entire glacial cycle. At the time when a full-size Cordilleran ice sheet is
absent, the Skeena ice cap appears to be fed by the eastwards precipitation
intrusion through a topographic window in the Coast Mountains. The ice cap acts
as a nucleation centre at the onset of the LGM readvance, and appears
consistent with the distinct glacial imprint of the Skeena Mountains landscape.

During deglaciation, none of the palaeo-temperature records used produces a
close agreement between the model results and the geological evidence. Although
the EPICA record yields a more realistic timing of the LGM and early
deglaciation, only the GRIP record produces a late-glacial readvance in areas
where it has been documented in the literature. Nonetheless, the
general patterns of deglaciation are consistent between both simulations driven
by the GRIP and EPICA records, and show a rapid deglaciation of the southern
half of the ice sheet, including a rapid northwards retreat across the Interior
Plateau of central British Columbia. This is followed by an opening of the ice
margin in the Liard Lowland, and a final retreat of the margin of the remaining
ice caps towards the Selwyn and, later, the Skeena mountains, which host the
last remnant of the ice sheet
during the early Holocene (10.9--9.5\,ka).

These results are strongly dependent on selected ice-sheet model (PISM),
surface
mass balance model (PDD) and climate forcing. Most importantly, our simplistic
palaeo-climate forcing does not include precipitation corrections in response
to the presence of an ice cover, potentially leading to overestimated glacial
extent and volume in continental regions. Nevertheless, our results identify
the Skeena Mountains as a key area to understanding
glacial dynamics of the Cordilleran ice sheet, highlighting the need for
further geological investigation of this region.

>>>>

However, on the Fraser Plateau, both
simulations predict negligible basal sliding during deglaciation
(Fig.~\ref{fig:lastflow}), but an impressive set of glacial lineations
indicate a substantial eastward flow component of the Cordilleran ice sheet
\citep{Prest.etal.1968, Kleman.etal.2010}.
The Fraser lineation set could thus present a smoking gun for the
reliability of the presented model results. One explanation for the incongruent
results could therefore be that missing feedback mechanisms between ice sheet
topography and regional climate position the modelled ice divide of the LGM ice
sheet too far to the east (Sect.~\ref{sec:mis2}; Fig.~\ref{fig:mis2};
\citealp{Seguinot.etal.2014}). A more westerly-located LGM ice divide would
certainly result in a different deglacial flow pattern over the Interior
Plateau. However, a more westerly-positioned LGM ice divide would certainly be
associated with a thinner ice sheet than that modelled here. Decreased ice
thickness would not promote warm-based conditions but, on the
opposite, enlarge the region of negligible basal sliding
(Fig.~\ref{fig:lastflow}). Thus, another explanation for the incongruent
results could be that the Fraser Plateau lineation system predates
deglaciation, and that deglacial landforms are largely absent. Alternatively,
local geothermal heat associated with volcanic activity on the Fraser Plateau
could have triggered the basal sliding
\citep[cf. Greenland ice sheet,][]{Fahnestock.etal.2001}.

@@

Numerical simulations of the Cordilleran ice sheet through the last glacial
cycle presented in this study consistently produce two glacial maxima during
MIS~4 (61.9--55.4\,ka, 3.8--8.7\,m~s.l.e) and MIS~2 (29.5--16.9\,ka,
8.24--8.66\,m~s.l.e.), two periods corresponding to documented extensive
glaciations. This result is independent of the palaeo-temperature record used
among the six selected for this study, and thus can be seen as a robust model
output, which broadly matches geological evidence. However, the timing of the
two glaciation peaks depends sensitively on which climate record is used to
drive the model. The timing of the LGM is best reproduced by the EPICA and
Vostok Antarctic ice core records, and occurs too early in all other
simulations. The mismatch is largest for the two northwest Pacific ODP
palaeo-temperature records, which are affected by the weakening of the
California current during the LGM.

In all simulations presented here, ice cover is limited to disjoint mountain
ice caps during most of the glacial cycle. The most persistent nucleation
centres are located in the Coast Mountains, the Columbia and Rocky mountains,
the Selwyn and MacKenzie mountains, and most importantly, in the Skeena
Mountains. Throughout the modelled last glacial cycle, the Skeena Mountains
host an ice cap which appears to be fed by moisture intruding inland from the
west through a topographic low in the Coast Mountains. The Skeena ice cap acts
as the main nucleation centre at the onset of the LGM readvance. As indicated
by persistent, warm-based ice in the model, this ice cap is perhaps associated
with the distinct glacial erosional imprint on the landscape of the Skeena
Mountains.

During deglaciation, none of the climate records used can be selected as
producing an optimal agreement between the model results and the geological
evidence. Although the EPICA-driven simulation yields the most realistic timing
of the LGM and, therefore, start of deglaciation, only the GRIP-driven
simulation produces late glacial readvances in areas where these have been
documented. Nonetheless, the patterns of ice sheet retreat are consistent
between the two simulations, and show a rapid deglaciation of the southern
sector of the ice sheet, including a rapid northwards retreat across the Fraser
Plateau of central British Columbia. The GRIP-driven simulation then produces a
late-glacial readvance of local ice caps in the Coast, Columbia and Rocky
mountains, and of the main body of the decaying Cordilleran ice sheet in the
Skeena and Selwyn mountains. In both simulations, this is followed by an
opening of the Liard Lowland, and a final retreat of the remaining ice caps
towards the Selwyn and, finally, the Skeena mountains, which hosts the last
remnant of the ice sheet during the early Holocene (10.9--9.5\,ka). Our results
identify the Skeena Mountains as a key area to understanding glacial dynamics
of the Cordilleran ice sheet, highlighting the need for further geological
investigation of this region.
